Self Referencing Sequences

### 2014-11-05 draft.tex
changed generating sets to generating cycles, made notation consistent

We present an algorithm that produces the unique self-referencing sequence given a generating cycle and a starting element. Assume the generating cycle is $(a_1,a_2,\dots,a_m)$ and that the starting element is some $a_i$. For notational simplicity, if we ever write down $a_k$ where $k > m$, we set this equal to $a_{k (\text{mod } m)}$.
\begin{quote}
\textbf{Algorithm:}\\
Suppose the first number of the sequence is $a_i$. Then the first block of the sequence must be a block of $a_i$ instances of $a_i$, so we extend the sequence to this block. If $a_i$ is not 1, then there is a second element of that block. We read that second element, and then we add a block of $a_i$ instances of $a_{i+1}$ to the end of that sequence. If $a_i$ is equal to 1, we write down $a_{i+1}$ in the second spot, and use the value of $a_{i+1}$ to dictate the length of the second block.\\

@@

We now introduce a notion which will help us categorize sequences as equivalent. 
\begin{definition} Consider a sequence $\{a_i\}$. We define the read distance for an index $i$ as the number of terms in between it $a_i$ and the beginning of the $i$th block of numbers.
\end{definition}
The read distance can be any nonnegative integer. For example if we consider the sequence $\{a_i\}$ generated by $(1,2,3)$ beginning with a 1  
\begin{align*}
1,2,2,3,3,1,1,1,2,2,2,3,1,2,3,3,1,1,2,2,3,3,3,\dots
\end{align*}
the read distance for $1$ is always $0$ since in a self-referencing sequence, the first number must refer to the first block. The read distance for $2$ is $0$ as the second block of numbers starts at the second number. We can continue to calculate the read distances for this sequence and see that they form a monotonically increasing sequence $0, 0 ,1,2,4,6,6,6,6,7,8,9,\ldots$. The read distance, can be thought of as the horizontal shift of the arrows in Figure~\ref{fig:onetwothree}.

@@

\begin{lemma} Consider a self referencing sequence $\{a_i\}$ generated by $(c_1, c_2, \ldots, c_n)$ with $m = \max(c_1, c_2, \ldots, c_n)$. Suppose the read distance for $i$ is $d$, then the read distance for $i+m$ is at least $d+1$. 
\end{lemma}
\begin{proof}
The read distances can stay constant from $i$ to $i+1$ only if there is a $a_{i+1} = 1$. There can be at most $m$ 1's in a contiguous block, so the read distance of $i+m$ is bigger than the read distance of $i$.  
\end{proof}

@@

We change focus in this section and consider how to generate self-referential sequences with an iterative process. We first look at the sequences generated by the cycle $C = (1,2)$.

@@

It turns out that we can be less restrictive about what our starting sequence, $s^{(0)}$, is. Going back to the case of $C = (1,2)$, we can let $s^{(0)}$ be \emph{any} sequence of 1's and 2's that starts with a 2. And still, the limiting sequence will be the self-referential sequence.
For example, suppose $s^{(0)} = 2,1,1,1,2$. The rules for generating $s^{(i)}$ give:
\begin{align*}
s^{(0)}&= \mathbf{2},1,1,1,2\\
s^{(1)} &= \mathbf{2,2,1},2,1,2,2\\
s^{(2)} &= \mathbf{2,2,1,1,2,1},1,2,1,1,2,2\\
s^{(3)} &= \mathbf{2,2,1,1,2,1,2,2,1,2},1,1,2,1,2,2,1,1\\
&\vdots
\end{align*}
Here, we bold the numbers that form a prefix of $s(C,2)$. We note that $s^\infty = s(C,a)$ because because $s^{(0)}$ contains a 2 at the beginning, which by the above theorem will produce $s(C,a)$ in the limit. We call the rest of each $s^{(i)}$ that is not a prefix of $s(C,2)$ the \emph{tail}.

@@

The iterative process given in Section~\ref{iterativeprocess} is one way to generate self-referencing sequences. However, for certain generating sets $A$, there is a simpler ways to perform this iteration using a fixed set of substitution rules. Take $A = \{1,3\}$ as an example, and consider the sequence that starts with a 3:
\begin{align*}
333111333131333111333\dots
\end{align*}
We can use the iterative process from Section~\ref{iterativeprocess}, and suppose that $s^{(0)} = 33$. We get:
\begin{align*}
\{s^{(0)}\} &= 3,3\\
\{s^{(1)}\} &= 3,3,3,1,1,1\\
\{s^{(2)}\} &= 3,3,3,1,1,1,3,3,3,1,3,1\\
\{s^{(3)}\} &= 3,3,3,1,1,1,3,3,3,1,3,1,3,3,3,1,1,1,3,3,3,1,3,3,3,1\\
&\cdots
\end{align*}
It turns out that we can model this evolution with the following substitution rules:
\begin{align*}
\text{Rule }A: 3,3 &\to 3,3,3,1,1,1 \\
\text{Rule }B: 3,1 &\to 3,3,3,1\\
\text{Rule }C: 1,1 &\to 3,1
\end{align*}
This is to be interpreted as follows. Starting from 3,3, the only possible rule to apply is Rule A, which gives $3,3,3,1,1,1$ at the following iteration. We then apply Rule A to $3,3$, Rule B to $3,1$ and Rule C to $1,1$. Note that at every step, we break up the sequence into chunks of 2 and then simultaneously apply \emph{all} the rules. This will always be possible, since the rules preserve the fact that these sequences have even length. So at the following step, we have 3,3,3,1,1,1,3,3,3,1,3,1, which is $s^{(2)}$. Notice that in general, after the $i$th application of the substitution rules, the sequence is equal to $s^{(i)}$.

@@

We would like to know if we can do this for any generating set $G$ and starting number $g$. In particular we would like to be able to simulate our original algorithm with a substitution rule. Previously we established that our algorithm had a state corresponding to 
\begin{itemize}
\item current block size
\item number of elements already written down in the current block
\item what number the current block is composed of 
\end{itemize}

When we apply our substitution rules we should not need to know what happened to blocks other than the one we are currently substituting for. Therefore, each rule must somehow encode the length of  blocks, the number of elements already written in the current block, and what number the current block is composed of. The first condition is simple to implement, as the rules will specify the length of blocks. In order to keep track of the numbers already written, each rule should be self-contained. That is each rule should write complete blocks. In order to comply with the third condition it is enough for the size of the generating set to divide the length of the rules. This is so that at the start of each rule, we always know which numbers to write.

Therefore, we can formulate a set of sufficient conditions to make rules to generate the self-referential sequence with generating set $G$. We want to have a set of rules $R$ where each rule $r \in R$ has length $|r| = l$, such that all $|G|^{l}$ strings are in the rule set and for each $r \in R$
\begin{itemize}
\item each rule writes blocks in the same order of $G$
\item $\sum_{i \in r} i$ is divisible by $l$
\item $l$ is divisible by $|G|$
\end{itemize}
Given these conditions on every rule, we can guarantee that the self-referential sequence is generated. Note that in this case, the above example does not satisfy our rules until we say that $1, 3 \to 3, 1, 1, 1$.

Another example, we can have $G = \{ 2, 4 \}$, and we can have that the rules be
\begin{align*}
\text{Rule }A: 2,2 &\to 2, 2, 4, 4 \\
\text{Rule }B: 2,4 &\to 2, 2, 4,4,4,4\\
\text{Rule }C: 4,2 &\to 2,2,2,2,4,4\\
\text{Rule }D: 4,4 &\to 2, 2, 2, 2, 4, 4, 4, 4
\end{align*}

One of the reasons why generating a rule set for $\{1, 2\}$ is complicated is that since $1, ..., 1$ must be in the rule set, the length of the rules must be even (since $|G| = |\{1, 2\}| = 2$). On the other hand, $2, 1, ..., 1$ must also be in the set and so the rule length must be odd.

@@

A topic of interest when considering a self referencing sequence is the relative density of the various elements of the generating set. It is unclear if such a density even exists and if so how to compute it for arbitrary generating sets. In this section, we show how to use substitution rules to compute relative densities for certain sequences if we assume such densities exists. We explain the method and compute the relative density of 3's for $G = \{ 1, 3\}$. We will also produce bounds on the density of $2$'s $G = \{ 1, 2\}$ in a way that suggests that the density converges to 2; however, this method is computationally difficult for smaller bounds since the number of cases to check increases exponentially.

To begin we will discuss a technique that can be used to compute the densities for some generating sets. In particular, suppose we have a generating set $G$ with a set of rules $R$ that satisfy the conditions from the section above. For each rule $r \in R$, we will say denote the rule as $r_l \rightarrow r_r$ as the inputs and the outputs.

@@

As discussed in the previous section, it is unclear how to establish a set of rules for $G = \{1 , 2\}$; however, there is some structure that we can exploit. In particular, we know that the sequences $1, 1, 1$ and $2, 2, 2$ will never appear in the self-referential sequence. This is because such sequences would be part of a block of size at least $3$.

@@

\begin{itemize}
\item In Section~\ref{equivalence}, we propose the idea that in some sense, the number that we start a sequence with does not matter. In other words, any sequence that uses a certain generating set is equivalent to any other sequence created by the same generating set.
\item In the same section, we propose a condition for determining when two sequences are not equivalent. We conjecture that if two sequences have unequal numbers with the same read distances, then they are not equivalent. It seems from trying cases that this condition is sufficient, but we have been unable to prove it.
\item In Section~\ref{iterativeprocess} we make a conjecture on when sequences are strongly and weakly convergent. We have not yet been able to investigate strong and weak convergence in great detail, as we have been unable to make precise mathematical statements about these types of convergence. Most of what we know about them comes from analyzing numerical data.
\item In Section~\ref{substitutionrules}, we look at substitution rules for certain ``nice" generating sets of numbers, such as $\{1,3\}$ and $\{2,4\}$. These sets allow for substitution rules because they satisfy the trio of properties we outline in that section. However, we do not know that these are the only properties. We leave unresolved the question of determining precisely which sets allow for substitution rules and which ones ones do not. The major difficulty here seems to come in proving that a certain generating set does \emph{not} allow for substitution rules.
\item In Section~\ref{density} we were able to compute the density of any sequence with generating set $\{1,3\}$, and we were able to give bounds on the density of any sequence with generating set $\{1,2\}$. However, we leave unresolved the question of what the density for $\{1,2\}$ actually is, or even if we can obtain tighter bounds.
\end{itemize}

>>>>

We present an algorithm that produces the unique self-referencing sequence given a generating cycle and a starting element. Assume the generating cycle is $(a_1 a_2 \dots a_m)$ and that the starting element is some $a_i$. For notational simplicity, if we ever write down $a_k$ where $k > m$, we set this equal to $a_{k (\text{mod } m)}$.
\begin{quote}
\textbf{Algorithm:}\\
Suppose the first number of the sequence is $a_i$. Then the first block of the sequence must be a block of $a_i$ instances of $a_i$, so we extend the sequence to this block. If $a_i$ is not 1, then there is a second element of that block. We read that second element, and then we add a block of $a_i$ instances of $a_{i+1}$ to the end of that sequence. If $a_i$ is equal to 1, we write down $a_{i+1}$ in the second spot, and use the value of $a_{i+1}$ to dictate the length of the second block.\\

@@

We now introduce a notion which will help us categorize sequences as equivalent. 
\begin{definition} Consider a sequence $\{a_i\}$. We define the read distance for an index $i$ as the number of terms in between it $a_i$ and the beginning of the $i$th block of numbers.
\end{definition}
The read distance can be any nonnegative integer. For example if we consider the sequence $\{a_i\}$ generated by $C = (1 2 3)$ beginning with a 1  
\begin{align*}
1,2,2,3,3,1,1,1,2,2,2,3,1,2,3,3,1,1,2,2,3,3,3,\dots
\end{align*}
the read distance for $1$ is always $0$ since in a self-referencing sequence, the first number must refer to the first block. The read distance for $2$ is $0$ as the second block of numbers starts at the second number. We can continue to calculate the read distances for this sequence and see that they form a monotonically increasing sequence $0, 0 ,1,2,4,6,6,6,6,7,8,9,\ldots$. The read distance, can be thought of as the horizontal shift of the arrows in Figure~\ref{fig:onetwothree}.

@@

\begin{lemma} Consider a self referencing sequence $\{a_i\}$ generated by $(c_1 c_2 \ldots c_n)$ with $m = \max(c_1, c_2, \ldots, c_n)$. Suppose the read distance for $i$ is $d$, then the read distance for $i+m$ is at least $d+1$. 
\end{lemma}
\begin{proof}
The read distances can stay constant from $i$ to $i+1$ only if there is a $a_{i+1} = 1$. There can be at most $m$ 1's in a contiguous block, so the read distance of $i+m$ is bigger than the read distance of $i$.  
\end{proof}

@@

We change focus in this section and consider how to generate self-referential sequences with an iterative process. We first look at the sequences generated by the cycle $C = (1 2)$.

@@

It turns out that we can be less restrictive about what our starting sequence, $s^{(0)}$, is. Going back to the case of $C = (1 2)$, we can let $s^{(0)}$ be \emph{any} sequence of 1's and 2's that starts with a 2. And still, the limiting sequence will be the self-referential sequence.
For example, suppose $s^{(0)} = 2,1,1,1,2$. The rules for generating $s^{(i)}$ give:
\begin{align*}
s^{(0)}&= \mathbf{2},1,1,1,2\\
s^{(1)} &= \mathbf{2,2,1},2,1,2,2\\
s^{(2)} &= \mathbf{2,2,1,1,2,1},1,2,1,1,2,2\\
s^{(3)} &= \mathbf{2,2,1,1,2,1,2,2,1,2},1,1,2,1,2,2,1,1\\
&\vdots
\end{align*}
Here, we bold the numbers that form a prefix of $s(C,2)$. We note that $s^\infty = s(C,a)$ because because $s^{(0)}$ contains a 2 at the beginning, which by the above theorem will produce $s(C,a)$ in the limit. We call the rest of each $s^{(i)}$ that is not a prefix of $s(C,2)$ the \emph{tail}.

@@

The iterative process given in Section~\ref{iterativeprocess} is one way to generate self-referencing sequences. However, for certain generating cycles $A$, there is a simpler ways to perform this iteration using a fixed set of substitution rules. Take $A = \{1,3\}$ as an example, and consider the sequence that starts with a 3:
\begin{align*}
333111333131333111333\dots
\end{align*}
We can use the iterative process from Section~\ref{iterativeprocess}, and suppose that $s^{(0)} = 33$. We get:
\begin{align*}
\{s^{(0)}\} &= 3,3\\
\{s^{(1)}\} &= 3,3,3,1,1,1\\
\{s^{(2)}\} &= 3,3,3,1,1,1,3,3,3,1,3,1\\
\{s^{(3)}\} &= 3,3,3,1,1,1,3,3,3,1,3,1,3,3,3,1,1,1,3,3,3,1,3,3,3,1\\
&\cdots
\end{align*}
It turns out that we can model this evolution with the following substitution rules:
\begin{align*}
\text{Rule }A: 3,3 &\to 3,3,3,1,1,1 \\
\text{Rule }B: 3,1 &\to 3,3,3,1\\
\text{Rule }C: 1,1 &\to 3,1
\end{align*}
This is to be interpreted as follows. Starting from 3,3, the only possible rule to apply is Rule A, which gives $3,3,3,1,1,1$ at the following iteration. We then apply Rule A to $3,3$, Rule B to $3,1$ and Rule C to $1,1$. Note that at every step, we break up the sequence into chunks of 2 and then simultaneously apply \emph{all} the rules. This will always be possible, since the rules preserve the fact that these sequences have even length. So at the following step, we have 3,3,3,1,1,1,3,3,3,1,3,1, which is $s^{(2)}$. Notice that in general, after the $i$th application of the substitution rules, the sequence is equal to $s^{(i)}$.

@@

We would like to know if we can do this for any generating cycle $G$ and starting number $g$. In particular we would like to be able to simulate our original algorithm with a substitution rule. Previously we established that our algorithm had a state corresponding to 
\begin{itemize}
\item current block size
\item number of elements already written down in the current block
\item what number the current block is composed of 
\end{itemize}

When we apply our substitution rules we should not need to know what happened to blocks other than the one we are currently substituting for. Therefore, each rule must somehow encode the length of  blocks, the number of elements already written in the current block, and what number the current block is composed of. The first condition is simple to implement, as the rules will specify the length of blocks. In order to keep track of the numbers already written, each rule should be self-contained. That is each rule should write complete blocks. In order to comply with the third condition it is enough for the size of the generating cycle to divide the length of the rules. This is so that at the start of each rule, we always know which numbers to write.

Therefore, we can formulate a set of sufficient conditions to make rules to generate the self-referential sequence with generating cycle $C$. We want to have a set of rules $R$ where each rule $r \in R$ has length $|r| = l$, such that all $|C|^{l}$ strings are in the rule set and for each $r \in R$
\begin{itemize}
\item each rule writes blocks in the same order of $C$
\item $\sum_{i \in r} i$ is divisible by $l$
\item $l$ is divisible by $|C|$
\end{itemize}
Given these conditions on every rule, we can guarantee that the self-referential sequence is generated. Note that in this case, the above example does not satisfy our rules until we say that $1, 3 \to 3, 1, 1, 1$.

Another example, we can have $C = (2 4)$, and we can have that the rules be
\begin{align*}
\text{Rule }A: 2,2 &\to 2, 2, 4, 4 \\
\text{Rule }B: 2,4 &\to 2, 2, 4,4,4,4\\
\text{Rule }C: 4,2 &\to 2,2,2,2,4,4\\
\text{Rule }D: 4,4 &\to 2, 2, 2, 2, 4, 4, 4, 4
\end{align*}

One of the reasons why generating a rule set for $(1 2)$ is complicated is that since $1, ..., 1$ must be in the rule set, the length of the rules must be even (since $|C| = 2$). On the other hand, $2, 1, ..., 1$ must also be in the set and so the rule length must be odd.

@@

A topic of interest when considering a self referencing sequence is the relative density of the various elements of the generating cycle. It is unclear if such a density even exists and if so how to compute it for arbitrary generating cycles. In this section, we show how to use substitution rules to compute relative densities for certain sequences if we assume such densities exists. We explain the method and compute the relative density of 3's for $C = (1 3)$. We will also produce bounds on the density of $2$'s $C = (1 2)$ in a way that suggests that the density converges to 2; however, this method is computationally difficult for smaller bounds since the number of cases to check increases exponentially.

To begin we will discuss a technique that can be used to compute the densities for some generating cycles. In particular, suppose we have a generating cycle $C$ with a set of rules $R$ that satisfy the conditions from the section above. For each rule $r \in R$, we will say denote the rule as $r_l \rightarrow r_r$ as the inputs and the outputs.

@@

As discussed in the previous section, it is unclear how to establish a set of rules for $C = (1 2)$; however, there is some structure that we can exploit. In particular, we know that the sequences $1, 1, 1$ and $2, 2, 2$ will never appear in the self-referential sequence. This is because such sequences would be part of a block of size at least $3$.

@@

\begin{itemize}
\item In Section~\ref{equivalence}, we propose the idea that in some sense, the number that we start a sequence with does not matter. In other words, any sequence that uses a certain generating cycle is equivalent to any other sequence created by the same generating cycle.
\item In the same section, we propose a condition for determining when two sequences are not equivalent. We conjecture that if two sequences have unequal numbers with the same read distances, then they are not equivalent. It seems from trying cases that this condition is sufficient, but we have been unable to prove it.
\item In Section~\ref{iterativeprocess} we make a conjecture on when sequences are strongly and weakly convergent. We have not yet been able to investigate strong and weak convergence in great detail, as we have been unable to make precise mathematical statements about these types of convergence. Most of what we know about them comes from analyzing numerical data.
\item In Section~\ref{substitutionrules}, we look at substitution rules for certain ``nice" generating cycles of numbers, such as $(13)$ and $(24)$. These sets allow for substitution rules because they satisfy the trio of properties we outline in that section. However, we do not know that these are the only properties. We leave unresolved the question of determining precisely which sets allow for substitution rules and which ones ones do not. The major difficulty here seems to come in proving that a certain generating cycle does \emph{not} allow for substitution rules.
\item In Section~\ref{density} we were able to compute the density of any sequence with generating cycle $(13)$, and we were able to give bounds on the density of any sequence with generating cycle $(12)$. However, we leave unresolved the question of what the density for $(12)$ actually is, or even if we can obtain tighter bounds.
\end{itemize}
